Verifying Cache Coherence Protocols for an Infinite Family of Cache Hierarchies

### 2014-02-06 paper/Background.tex
don't like relying on fixity except for -> operator

Each input to the memory subsystem is isomorphic to a function
$\reqFn : Processor \times \mathbb{N} @-> \Request$, where $\reqFn(p, i)$ models the
$i^{th}$ request issued by processor $p$. An element of the set \Request{} (the
range of \reqFn{}) contains fields (a) address (\addrQ) corresponding to
the request, (b) \desc{} indicating whether the request is a load or a store,
and (c) data (\dataQ) containing the data supplied by the processor for stores.

>>>>

Each input to the memory subsystem is isomorphic to a function
$\reqFn : (Processor \times \mathbb{N}) @-> \Request$, where $\reqFn(p, i)$ models the
$i^{th}$ request issued by processor $p$. An element of the set \Request{} (the
range of \reqFn{}) contains fields (a) address (\addrQ) corresponding to
the request, (b) \desc{} indicating whether the request is a load or a store,
and (c) data (\dataQ) containing the data supplied by the processor for stores.

### 2014-02-06 paper/System.tex
don't like relying on fixity except for -> operator

Each node contains the fields mentioned in Figure~\ref{hier}. We will
explain what the fields mean in the following discussion.

@@

Each node maintains a permission state for each address $a$, $c.\state[a]$, in
addition to the data associated with the address, $c.\data[a]$. The permission
state denotes which operations may be performed on the data. An address may be
in the \emph{read-write} state \Mo{} (for ``modifiable''),
the \emph{read-only} state \Sh{} (for ``shared''), or in the \emph{invalid} state
\In{} (indicating that it may be neither read nor written). The options form a natural
total order of permissions ($\In < \Sh < \Mo$). For a leaf cache, the meaning
of these permission states is straightforward: a load request may be processed
only if the leaf is at least in the \Sh{} state, and a store request may be
processed only if it is in the \Mo{} state. The state of a non-leaf cache indicates
an \emph{upper bound} on the permission level of any cache in the associated subtree.
If a cache does not include a particular location,
we consider that address to have permission \In.  Initially, all addresses in
each cache are in the \In{} state, save for those in the root cache (\ie{} main memory),
which are in the \Mo{} state.

@@

In addition to the permission states, each cache also maintains a
temporary \emph{wait state} for each address to deal with transitioning between
permission levels. The wait state consists of two parts: (a) a Boolean
representing whether the cache has sent a request to its parent cache earlier
and is waiting for a response (denoted by $c.\wait[a]$ for address $a$) and (b)
one of the three values \Mo, \Sh{} or \In{} representing the state upgrade that
the cache is waiting to make (denoted by $c.\waitS[a]$ for address $a$).
Similarly, each non-leaf cache has the corresponding wait states for its
directory.  $p.\dwait[c][a]$ denotes that cache $p$ is waiting for a response
from its child $c$ for address $a$, and $p.\dwaitS[c][a]$ denotes the state to
which $p$ is waiting for $c$ to downgrade. Both \wait{} and \dwait{} are
\False{} initially for all caches and addresses,
because no cache is waiting for any response from either its child or its
parent.

The behavior of the system is described via the atomic state transitions, which
are effectively a nondeterministic operational semantics. These transitions
have two parts: (a) the guard dictating when the transition can fire, and (b)
the action defining how the state of the system changes when the transition
happens. Logically, these transitions happen one at a time.  Figure~\ref{trans}
gives the atomic transitions for the MSI protocol.  The actions in the
transitions in Figure~\ref{trans} show only the specific states that change,
and other state components are assumed to remain the same.  The notation ($x
\Leftarrow y$;) in a transition indicates setting of a specific state $x$ of
the overall system to value $y$.

@@

Figure~\ref{childside} shows the transitions that result in an upgrade of the
state of a cache. It involves a cache sending an upgrade request to the its
parent if the cache is not already waiting for some response for that address
from the parent (ChildSendReq). This request is received by the parent
(ParentRecvReq). If the directory entries for all of the other children are
\emph{compatible} with the upgrade request, and the parent's permission state has
at least the requested value, it sends an upgrade response to the cache, while
changing the appropriate directory state. The notation \compat{} used in
the ParentRecvReq transition of Figure~\ref{trans} is defined as follows:
\begin{multline*}
\small
\compat(p, c, a, s) = \forall i \ne c. \;
\mylet s_i := p.\dstate[i][a] \myin\\
 (s = \Mo @-> s_i = \In) \wedge (s = \Sh @-> s_i = \Sh) \wedge (s = \In @-> s_i = \Mo)
\end{multline*}
Finally, the child receives the response from the parent (ChildRecvResp)
and sets its state, resetting its wait state if the response contains the
desired upgrade.

@@

Because of the presence of voluntary responses, a child could have already
downgraded to a state required by a downgrade request from its parent. In such
cases, the child simply drops the downgrade request (ChildDropReq).

@@

Note that the atomic transitions in Figure~\ref{trans} access only local states,
\ie{} a transition can read or write states only corresponding to a single cache
and/or the channel the cache is connected to. This restriction allows such a
system to be implemented directly into hardware. Bluespec SystemVerilog (BSV),
for instance, directly converts these transitions into efficient synchronous
hardware. Though these transitions logically happen one by one, BSV executes
several transitions simultaneously, as long as they do not access the same
state, \ie{} do not \emph{conflict}~\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
Sometimes, these transitions (both the guards and the actions) may not be
amenable to single-cycle implementations in hardware. For example, writing a
register array can actually take several hardware clock cycles, though
logically it is a single transition. Karczmarek \etal \cite{Karczmarek} has
provided a scheme to convert atomic transitions into synchronous hardware in
which each transition can span several clock cycles.

>>>>

Each node contains the fields mentioned in Figure \ref{hier}. We will
explain what the fields mean in the following discussion.

@@

Each node maintains a permission state for each address $a$, $c.\state[a]$, in
addition to the data associated with the address, $c.\data[a]$. The permission
state denotes which operations may be performed on the data. An address may be
in the \emph{read-write} state \Mo{} (for ``modifiable''),
the \emph{read-only} state \Sh{} (for ``shared''), or in the \emph{invalid} state
\In{} (indicating that it may be neither read nor written). The options form a natural
total order of permissions ($\In < \Sh < \Mo$). For a leaf cache, the meaning
of these permission states is straightforward: a load request may be processed
only if the leaf is at least in the \Sh{} state, and a store request may be
processed only if it is in the \Mo{} state. The state of a non-leaf cache indicates
an \emph{upper bound} on the permission level of any cache in the associated subtree.
If a cache does not include a particular location,
we consider that address to have permission \In.  Initially, all addresses in
each cache are in the \In{} state, save for those in the root cache (i.e., main memory),
which are in the \Mo{} state.

@@

In addition to the actual permission states, each cache also maintains a
temporary \emph{wait state} for each address to deal with transitioning between
permission levels. The wait state consists of two parts: (a) a Boolean
representing whether the cache has sent a request to its parent cache earlier
and is waiting for a response (denoted by $c.\wait[a]$ for address $a$) and (b)
one of the three values \Mo, \Sh{} or \In{} representing the state upgrade that
the cache is waiting to make (denoted by $c.\waitS[a]$ for address $a$).
Similarly, each non-leaf cache has the corresponding wait states for its
directory.  $p.\dwait[c][a]$ denotes that cache $p$ is waiting for a response
from its child $c$ for address $a$, and $p.\dwaitS[c][a]$ denotes the state to
which $p$ is waiting for $c$ to downgrade. Both \wait{} and \dwait{} are
\False{} initially for all caches and addresses,
because no cache is waiting for any response from either its child or its
parent initially.

The behavior of the system is described via the atomic state transitions, which
are effectively a nondeterministic operational semantics. These transitions
have two parts: (a) the guard dictating when the transition can fire, and (b)
the action defining how the state of the system changes when the transition
happens. Logically, these transitions happen one at a time.  Figure~\ref{trans}
gives the atomic transitions for the MSI protocol.  The actions in the
transitions in Figure \ref{trans} show only the specific states that change,
and other state components are assumed to remain the same.  The notation ($x
\Leftarrow y$;) in a transition indicates setting of a specific state $x$ of
the overall system to value $y$.

@@

Figure~\ref{childside} shows the transitions that result in an upgrade of the
state of a cache. It involves a cache sending an upgrade request to the its
parent if the cache is not already waiting for some response for that address
from the parent (ChildSendReq). This request is received by the parent
(ParentRecvReq). If the directory entries for all of the other children are
\emph{compatible} with the upgrade request, and the parent's permission state has
at least the requested value, it sends an upgrade response to the cache, while
changing the appropriate directory state. The notation \compat{} used in
the ParentRecvReq transition of Figure \ref{trans} is defined as follows:
\begin{multline*}
\small
\compat(p, c, a, s) = \forall i \ne c. \;
\mylet s_i := p.\dstate[i][a] \myin\\
 (s = \Mo @-> s_i = \In) \wedge (s = \Sh @-> s_i = \Sh) \wedge (s = \In @-> s_i = \Mo)
\end{multline*}
Finally, the child cache receives the response from the parent (ChildRecvResp)
and sets its state while resetting its wait state if the response contains the
desired upgrade.

@@

Because of the presence of voluntary responses, a child could have already
downgraded to a state required by a downgrade request from its parent. In such
a case, the child simply drops the downgrade request (ChildDropReq).

@@

Note that the atomic transitions in Figure~\ref{trans} access only local states,
\ie{} a transition can read or write states only corresponding to a single cache
and/or the channel the cache is connected to. This restriction allows such a
system to be implemented directly into hardware. Bluespec System Verilog (BSV),
for instance, directly converts these transitions into efficient synchronous
hardware. Though these transitions logically happen one by one, BSV executes
several transitions simultaneously, as long as they do not access the same
state, \ie{} do not \emph{conflict}~\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
Sometimes, these transitions (both the guards and the actions) may not be
amenable to single-cycle implementations in hardware. For example, writing a
register array can actually take several hardware clock cycles, though
logically it is a single transition. Karczmarek \etal~\cite{Karczmarek} have
provided a scheme to convert atomic transitions into synchronous hardware in
which each transition can span several clock cycles.
